---
title: "Dengue Forecasting"
author: "Axel Forveille, MAI Tien Long"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r echo=FALSE, message=FALSE}
##### Start an R session
rm(list=objects())
```
## Introduction 
The purpose of the project is to predict the number of dengue cases in two cities, San Juan and Iquitos for each week, using environmental and climate variables. The challenge was organized in 2015 by several departments in the U.S. Federal Government (Department of Health and Human Services, Department of Defense, Department of Commerce, and the Department of Homeland Security), with the support of the Pandemic Prediction and Forecasting Science and Technology Interagency Working Group under the National Science and Technology Council (https://dengueforecasting.noaa.gov/).

#### Data 
The data for the challenge consists of:   

* City and week indicators.   
* NOAA's GHCN daily climate data weather station measurements.  
* PERSIANN satellite precipitation measurements.
* NOAA's NCEP Climate Forecast System Reanalysis measurements.
* Satellite vegetation. 
* Prediction for the number of cases for each week.

Additionally, we downloaded the social and economic data from WorldBank and we chose several parameters that might explain the number of cases:

* Total population       
* population_density_people_per_sq_km_of_land_area       
* forest_area_sq_km       
* gdp_current_us        
* employment_to_population_average      
* Population age percentage: 0 - 9, 10 - 20, 20 - 60, 60+            

Another data that can affect the spread of the disease is the number of migration, however, we couldn't find the data.

#### Methods

##### Nagative Binomial Distribution model
As the total number of cases does not follow a Gaussian distribution, and this is a counting variable, we should use the Poisson distribution or the Negative Binomial distribution model. 

##### Model GAM 

##### Regression Tree 

##### Random Forest

##### Times Series

##### Gradient Boosting

## Data Wrangling and Exploration 

## Modeling and prevision

##### Nagative Binomial Distribution model

##### Model GAM 

##### Regression Tree 

##### Random Forest

##### Times Series (ARIMA/SARIMA)

##### Gradient Boosting

## Conclusion


